2024 Big Data Bowl Submission
Names: Michael Labbe, Gabriel Tauro, David Smiley, and Alejandro Pesantes
Focus: Predicting linebacker block shed success and providing final rank out

```{r}
# Load the required libraries
library(dplyr)
library(tidyr)
library(readxl)
library(lubridate)
```


```{r}
# Load the dataset
data <- read.csv("BDB_all_weeks.csv") 
nrow(data)
```


```{r}
# Filter rows where playDescription does not contain "up the middle"
data <- data %>%
  filter(!grepl("up the middle", playDescription, ignore.case = TRUE))
nrow(data)
```


```{r}
# Filter rows where playDescription does not contain "TOUCHDOWN"
data <- data %>%
  filter(!grepl("TOUCHDOWN", playDescription))
nrow(data)
```


```{r}
# Filter rows where playDescription does not contain "No Play"
data <- data %>%
  filter(!grepl("No Play", playDescription))
nrow(data)

# Read in the "no_plays.csv" dataset
no_plays <- read.csv("no_plays.csv") 

# Assuming "data" and "no_plays" are your existing and new dataframes
data <- merge(data, no_plays, by = c("gameId", "playId"), all.x = TRUE)

# Identify rows to be removed
rows_to_remove <- !complete.cases(data)

# Remove rows from the original "data" dataframe
data <- data[rows_to_remove, ]

# Print the updated "data" dataframe
nrow(data)
```


```{r}
# Filter rows where final_position is not "TE"
data <- data %>% filter(final_position != "TE")
nrow(data)
```


```{r}
# Function to convert height to inches
convert_height_to_inches <- function(height_str) {
  parts <- strsplit(height_str, "-")[[1]]
  feet <- as.numeric(parts[1])
  inches <- as.numeric(parts[2])
  total_inches <- feet * 12 + inches
  return(total_inches)
}
```


```{r}
# Apply the function to the height column
data$height <- sapply(data$height, convert_height_to_inches)
```


```{r}
# Creating a def vs offense difference
data <- data %>%
  mutate(defPos = ifelse(final_position %in% c("MLB", "ILB", "OLB"), 1, 0))
nrow(data)
```


```{r}
# Define a function to calculate the distance between two points (x1, y1) and (x2, y2)
distance <- function(x1, y1, x2, y2) {
  sqrt((x1 - x2)^2 + (y1 - y2)^2)
}
```


```{r}
# Create a new data frame that pairs up subjects with defPos 1 and 0
data <- data %>%
  filter(defPos %in% c(0, 1)) %>%
  left_join(data %>% filter(defPos == 1), by = c("gameId", "playId", "time"), suffix = c("", "_Linebacker")) %>%
  filter(defPos == 0 & defPos_Linebacker == 1) %>%
  mutate(dist = distance(x, y, x_Linebacker, y_Linebacker)) %>%
  filter(dist <= 1)
nrow(data)
```


```{r}
close_pairs <- data

# Get the Height Difference
close_pairs$Height_Difference <- close_pairs$height_Linebacker - close_pairs$height

# Get the Weight Difference
close_pairs$Weight_Difference <- close_pairs$weight_Linebacker - close_pairs$weight

# Get the Speed Difference
close_pairs$Speed_Difference <- close_pairs$s_Linebacker - close_pairs$s

# Get the Acceleration Difference
close_pairs$Acceleration_Difference <- close_pairs$a_Linebacker - close_pairs$a

# Get rid of plays where the speed difference exceeded 2 for more than half of the time
close_pairs <- close_pairs %>%
  group_by(gameId, playId) %>%
  filter(mean(abs(Speed_Difference) > 2, na.rm = TRUE) <= 0.5)

nrow(close_pairs)
```


```{r}
# Convert the relevant columns to numeric (blocker orientation)
close_pairs$o <- as.numeric(close_pairs$o)
close_pairs$o_Linebacker <- as.numeric(close_pairs$o_Linebacker)
```


```{r}
# Calculate the angular difference
calculate_angular_difference <- function(angle1, angle2) {
  diff <- abs(angle1 - angle2) %% 360
  min_diff <- ifelse(diff > 180, 360 - diff, diff)
  return(min_diff)
}

# Apply the function to create a new column
close_pairs$Angular_Difference <- calculate_angular_difference(close_pairs$o, close_pairs$o_Linebacker)

# Calculate the scaled difference
close_pairs$Scaled_Angular_Difference <- abs(close_pairs$Angular_Difference) / 180

# Use subset function to filter rows
close_pairs <- subset(close_pairs, Scaled_Angular_Difference >= 0.6)

nrow(close_pairs)
```


```{r}
# Load the tackles dataset
tackle_data <- read.csv("tackles.csv")
```


```{r}
# Rename the column
tackle_data <- tackle_data %>%
  rename(nflId_Linebacker = nflId)
```


```{r}
# Merge the dataframes by playId, gameId, and nflId_1
merged_data <- merge(close_pairs, tackle_data, by = c("playId", "gameId", "nflId_Linebacker"), all.x = TRUE)
```


```{r}
# Fill missing values in the added columns with zeros
merged_data <- merged_data %>%
  mutate(across(colnames(tackle_data)[-c(1:3)], ~ifelse(is.na(.), 0, .)))
```


```{r}
# Rename data set
close_pairs <- merged_data
head(close_pairs, 5)
nrow(close_pairs)
```


```{r}
# Create the final dataset structure
result <- close_pairs %>%
  group_by(gameId, playId, nflId, nflId_Linebacker, displayName) %>%
  summarize(
    playId = playId,
    nflId_Blocker = first(nflId),
    nflId_Linebacker = first(nflId_Linebacker),
    time = time,
    dist = dist,
    Blocker_x = x,
    Blocker_y = y,
    Linebacker_x = x_Linebacker,
    Linebacker_y = y_Linebacker,
    possessionTeam = first(possessionTeam),
    defensiveTeam = first(defensiveTeam_Linebacker),
    playResult = first(playResult),
    goodRunDefense = first(goodRunDefense),
    week = week,
    homeFinalScore = first(homeFinalScore),
    visitorFinalScore = first(visitorFinalScore),
    Frames = n_distinct(frameId),
    Time_Spent_Engaged = Frames / 10,
    Two_Yards_Or_Less = first(goodRunDefense),
    Blocker_Name = first(displayName),
    Blocker_Position = first(final_position),
    Blocker_Speed = s,
    Blocker_Acceleration = a,
    Blocker_Orientation = o,
    Blocker_Height = first(height),
    Blocker_Weight = first(weight),
    Blocker_BirthDate = first(birthDate),
    Blocker_College = first(collegeName),
    Blocker_Direction = dir,
    Linebacker_Name = first(displayName_Linebacker),
    Linebacker_Position = first(final_position_Linebacker),
    Linebacker_Speed = s_Linebacker,
    Linebacker_Acceleration = a_Linebacker,
    Linebacker_Orientation = o_Linebacker,
    Linebacker_Height = first(height_Linebacker),
    Linebacker_Weight = first(weight_Linebacker),
    Linebacker_BirthDate = first(birthDate_Linebacker),
    Linebacker_College = first(collegeName_Linebacker),
    Linebacker_Direction = dir_Linebacker,
    Average_Direction_Degrees = mean(atan2(Linebacker_y, Linebacker_x), na.rm = TRUE) * (180 / pi),
    Acceleration_Difference = Acceleration_Difference,
    Height_Difference = Height_Difference,
    Weight_Difference = Weight_Difference,
    Week = week,
    Speed_Difference = Speed_Difference,
    Squareness = Scaled_Angular_Difference,
    tackle = tackle,
    assist = assist,
    forcedFumble = forcedFumble,
    pff_missedTackle = pff_missedTackle,
    .groups = 'keep'
  )
nrow(result)
```



```{r}
# Filter out engagements lasting longer than 2.5 seconds (at that point, they're just standing in vicinity)
result2 <- result %>%
  filter(Time_Spent_Engaged <= 2.5)

nrow(result2)
  
result2 <- result2 %>%
  filter(Time_Spent_Engaged >= 0.25)

nrow(result2)

 result2 <- result2 %>%
   group_by(gameId, playId, nflId_Linebacker) %>%
   arrange(row_number()) %>%
   mutate(
     row_count = n(),
     middle_row = ifelse(row_count %% 2 == 0, row_count %/% 2 + 1, (row_count + 1) %/% 2),
     is_middle_row = row_number() == middle_row
   ) %>%
   filter(is_middle_row | row_number() == 1) %>%
   select(-row_count, -middle_row)%>%
   ungroup()
 nrow(result2)
 
 library(dplyr)

# Assuming your dataframe is named df
result2 <- result2 %>%
  group_by(gameId, playId, nflId_Linebacker) %>%
  mutate(Change_Linebacker_Direction = abs(diff(Linebacker_Direction)),
         Change_Linebacker_Speed = diff(Linebacker_Speed),
         Change_Linebacker_Acceleration = diff(Linebacker_Acceleration),
         Change_Linebacker_Orientation = abs(diff(Linebacker_Orientation)),
         Change_Blocker_Direction = abs(diff(Blocker_Direction)),
         Change_Blocker_Speed = diff(Blocker_Speed),
         Change_Blocker_Acceleration = diff(Blocker_Acceleration),
         Change_Blocker_Orientation = abs(diff(Blocker_Orientation)))

# Display the first few rows of the dataframe
nrow(result2)

# Assuming df is your dataframe
#result2 <- result2[duplicated(result2[c("gameId", "playId", "nflId")]) | #!duplicated(result2[c("gameId", "playId", "nflId")], fromLast = TRUE), ]

library(dplyr)

# Assuming your dataframe is named "your_dataframe"

result2 <- result2 %>%
  group_by(gameId, playId, nflId_Linebacker) %>%
  filter(row_number() == 1)

# Now, "filtered_dataframe" contains only the 1st occurrence of each unique playId, gameId, and nflId combination


nrow(result2)
```


```{r}
# Create linebacker data set
result_Linebacker <- result2 %>%
  group_by(Linebacker_Name) %>%
  filter(n_distinct(paste(playId, gameId, nflId_Linebacker)) >= 30) %>%
  ungroup()
```


```{r}
# Create lineman data set
result_Lineman <- result2 %>%
  group_by(Blocker_Name) %>%
  filter(n_distinct(paste(playId, gameId, nflId_Blocker)) >= 30) %>%
  ungroup()
nrow(result_Linebacker)
nrow(result_Lineman)
```


```{r}
write.csv(result_Linebacker, "~\\NFL Big Data Bowl\\nfl-big-data-bowl-2024\\BDB_Linebackers.csv", row.names=FALSE)
write.csv(result_Lineman, "~\\NFL Big Data Bowl\\nfl-big-data-bowl-2024\\BDB_Lineman.csv", row.names=FALSE)
```


```{r}
library(tidyverse)
library(xgboost)
library(shapr)
library(SHAPforxgboost)
library(caret)
library(pROC)

linebacker_data = read.csv("BDB_Linebackers.csv") %>%
  select(c(goodRunDefense, Acceleration_Difference, Height_Difference,
           Weight_Difference, Speed_Difference, Squareness, Time_Spent_Engaged, Linebacker_Speed, Linebacker_Acceleration, Linebacker_Height, Linebacker_Weight, Linebacker_BirthDate, Change_Linebacker_Direction, Change_Linebacker_Speed, Change_Linebacker_Acceleration, Blocker_Speed, Blocker_Acceleration, Blocker_Height, Blocker_Weight, Change_Blocker_Direction, Change_Blocker_Speed, Change_Blocker_Acceleration))

linebacker_data3 = read.csv("BDB_Linebackers.csv") %>%
  select(c(goodRunDefense, Acceleration_Difference, Height_Difference,
           Weight_Difference, Speed_Difference, Squareness, Time_Spent_Engaged, Linebacker_Speed, Linebacker_Acceleration, Linebacker_Height, Linebacker_Weight, Linebacker_BirthDate, Average_Direction_Degrees, defensiveTeam, Week))
```


```{r}
# Calculate the average seconds for each defensiveTeam and week
avg_seconds <- linebacker_data3 %>%
  group_by(defensiveTeam, Week) %>%
  summarize(avg_seconds = mean(Time_Spent_Engaged))

# Rank the average seconds within each week
avg_seconds <- avg_seconds %>%
  group_by(Week) %>%
  mutate(rank = rank(avg_seconds)) %>% 
  arrange(desc(avg_seconds))

# Print the result
print(avg_seconds)

# Calculate overall average avg_seconds for each defensive team
overall_avg <- avg_seconds %>%
  group_by(defensiveTeam) %>%
  summarize(overall_avg_seconds = mean(avg_seconds))

# Identify the top 5 most consistent teams
top_consistent_teams <- overall_avg %>% 
  top_n(6, wt = -overall_avg_seconds) %>%
  pull(defensiveTeam)

# Identify the top 5 most consistent teams
least_consistent_teams <- overall_avg %>% 
  top_n(5, wt = overall_avg_seconds) %>%
  pull(defensiveTeam)

print(least_consistent_teams)
print(top_consistent_teams)

library(ggplot2)
library(dplyr)

# Define custom colors for each team
custom_colors <- c("BUF" = "#C60C30", "TB" = "#FF7900", "IND" = "#002C5F", "NYJ" = "#125740", "PHI" = "#A5ACAF")

# Filter data for selected teams
selected_teams <- c("BUF", "TB", "IND", "NYJ", "PHI")
filt_data <- avg_seconds[avg_seconds$defensiveTeam %in% selected_teams, ]

# Filter data for Week 1 and 9
image_data <- filt_data %>%
  filter(Week %in% c(1, 9))

# Create a ggplot
ggplot(filt_data, aes(x = as.factor(Week), y = avg_seconds, group = defensiveTeam)) +
  
  # Connect data points with lines for each defensive team
  geom_line(aes(color = defensiveTeam), size = 5, position = position_dodge(width = 0.7)) +
  
  # Add images for each defensive team
  geom_image(data = image_data, aes(x = as.factor(Week), y = avg_seconds, image = paste0(trimws(defensiveTeam), ".png")), size = 0.08, position = position_dodge(width = 0.7)) +
  
  # Add text labels for each data point
  geom_text(aes(label = sprintf("%.2f", avg_seconds), color = defensiveTeam),
            position = position_dodge(width = 0.7),
            vjust = -1.5, hjust = 0.5, size = 7) +
  
  # Define custom colors
  scale_color_manual(values = custom_colors) +
  
  # Set plot space at the top
  scale_y_continuous(expand = c(0.05, 0.1)) +
  
  # Remove the legend for the color aesthetic
  guides(color = FALSE) +
  
  # Add labels and titles
  labs(
    title = "Best Defensive Teams in Overall Average Time Spent Engaged Week by Week",
    x = "Week",
    y = "Average Seconds"
  ) +
  
  # Theme settings
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))






# Define custom colors for each team
custom_colors <- c("ARI" = "#C41E3A", "LAC" = "#0080C6", "SF" = "#B3995D", "NE" = "#002244", "PIT" = "#FFB612")

# Filter data for selected teams
selected_teams <- c("ARI", "LAC", "SF", "NE", "PIT")
filtered_data <- avg_seconds[avg_seconds$defensiveTeam %in% selected_teams, ]

# Create a ggplot
ggplot(filtered_data, aes(x = as.factor(Week), y = avg_seconds, group = defensiveTeam)) +
  
  # Connect data points with lines for each defensive team
  geom_line(aes(color = defensiveTeam), size = 5, position = position_dodge(width = 0.7)) +
  
  # Add images for each defensive team only for Week 1, 8 (SF and PIT)
  geom_image(data = subset(filtered_data, Week %in% c(8) & (defensiveTeam == "SF" | defensiveTeam == "PIT")),
             aes(x = as.factor(Week), y = avg_seconds, image = paste0(trimws(defensiveTeam), ".png")),
             size = 0.08, position = position_dodge(width = 0.7)) +
  
  # Add images for each defensive team only for Week 1 and 9
  geom_image(data = subset(filtered_data, Week %in% c(1, 9)),
             aes(x = as.factor(Week), y = avg_seconds, image = paste0(trimws(defensiveTeam), ".png")),
             size = 0.08, position = position_dodge(width = 0.7)) +
  
  # Add text labels for each data point
  geom_text(aes(label = sprintf("%.2f", avg_seconds), color = defensiveTeam),
            position = position_dodge(width = 0.7),
            vjust = -0.7, hjust = 0.5, size = 8) +
  
  # Define custom colors
  scale_color_manual(values = custom_colors) +
  
  # Set plot space at the top
  scale_y_continuous(expand = c(0.05, 0.1)) +
  
  # Remove the legend for the color aesthetic
  guides(color = FALSE) +
  
  # Add labels and titles
  labs(
    title = "Worst Defensive Teams in Overall Average Time Spent Engaged Week by Week",
    x = "Week",
    y = "Average Seconds"
  ) +
  
  # Theme settings
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
colnames(read.csv("BDB_Linebackers.csv"))

# Install and load the lubridate package
# install.packages("lubridate")
library(lubridate)

# Assuming 'linebacker_data' is your data frame and 'Linebacker_BirthDate' is the column containing birthdates
linebacker_data$Linebacker_BirthDate <- mdy(linebacker_data$Linebacker_BirthDate)

# Set the end date
end_date <- as.Date("2022-12-01")

# Calculate the age
linebacker_data$Linebacker_Age <- interval(linebacker_data$Linebacker_BirthDate, end_date) %/% years(1)

# Remove the 'Linebacker_BirthDate' column
linebacker_data <- linebacker_data %>%
  select(-Linebacker_BirthDate)
```


### goodRunDefense Response
```{r}
set.seed(220)

# Split the data into training and testing sets
train_indices <- sample(c(TRUE, FALSE), nrow(linebacker_data), replace=TRUE, prob=c(0.8,0.2))
train <- linebacker_data[train_indices, ]
test <- linebacker_data[-train_indices, ]

X_train = as.matrix(train %>% select(-goodRunDefense))
y_train = as.matrix(train %>% select(goodRunDefense))
X_test = as.matrix(test %>% select(-goodRunDefense))
y_test = as.matrix(test %>% select(goodRunDefense))

nrow(train)
nrow(test)

# Train XGBoost model
model <- xgboost(
  data = as.matrix(X_train),
  label = y_train,
  nrounds = 1000,
  early_stopping_rounds = 20,
  objective = "binary:logistic",
  scale_pos_weight = sum(y_train == 0) / sum(y_train == 1),
  verbose = FALSE
  #params = list(min_child_weight = 4)
)

# Make predictions on the test set
predictions <- predict(model, as.matrix(X_test))

# Ensure both vectors have the same levels
levels_y <- levels(factor(y_test))
levels_predictions <- levels(factor(round(predictions)))

# Match levels
y_test <- factor(y_test, levels = levels_y)
predictions <- factor(round(predictions), levels = levels_predictions)

# Explain the model using shapr
shap_values <- shap.values(model, X_train)

# Plot feature importance based on SHAP values
print("Mean SHAP Scores:")
print(shap_values$mean_shap_score)

# Evaluate model performance
accuracy <- sum(predictions == y_test) / length(y_test)
print(paste("Accuracy:", round(accuracy * 100, 2), "%"))

# Additional performance metrics
conf_matrix <- confusionMatrix(predictions, y_test)
print("Confusion Matrix:")
print(conf_matrix)
```



```{r}
# To prepare the long-format data:
shap_long <- shap.prep(xgb_model = model, X_train = X_train)
# is the same as: using given shap_contrib
shap_long <- shap.prep(shap_contrib = shap_values$shap_score, X_train = X_train)
shap.plot.summary(shap_long)

summary(shap_long)
```



```{r}
# Load the dataset
data_LBs <- read.csv("BDB_Linebackers.csv") 
nrow(data_LBs)
head(data_LBs, 5)

library(lubridate)

# Assuming 'linebacker_data' is your data frame and 'Linebacker_BirthDate' is the column containing birthdates
data_LBs$Linebacker_BirthDate <- mdy(data_LBs$Linebacker_BirthDate)

# Set the end date
end_date <- as.Date("2022-12-01")

# Calculate the age
data_LBs$Linebacker_Age <- interval(data_LBs$Linebacker_BirthDate, end_date) %/% years(1)

# Remove the 'Linebacker_BirthDate' column
data_LBs <- data_LBs %>%
  select(-Linebacker_BirthDate)

```


# Try a GLM model:
```{r}
set.seed(1842)

# Fit a linear regression (GLM) model
selected_variables <- c("Speed_Difference", "Acceleration_Difference", "Squareness",  "Linebacker_Speed", "Linebacker_Acceleration", "Linebacker_Height", "Linebacker_Age", "Height_Difference", "Change_Linebacker_Direction", "Change_Linebacker_Orientation", "Change_Linebacker_Speed", "Change_Linebacker_Acceleration", "Blocker_Speed", "Blocker_Acceleration", "Blocker_Height", "Blocker_Weight", "Change_Blocker_Direction", "Change_Blocker_Speed", "Change_Blocker_Acceleration")

glm_model <- glm(Time_Spent_Engaged ~ ., data = data_LBs[, c(selected_variables, "Time_Spent_Engaged")], family = "gaussian")

# Output the summary
summary(glm_model)

glm_model$deviance

#calculate McFadden's R-squared for model
with(summary(glm_model), 1 - deviance/null.deviance)
```


# Next, let's try XGBoost:
```{r}
library(xgboost)

# Set seed for reproducibility
set.seed(1842)

# Split the data into training and testing sets
train_indices <- sample(c(TRUE, FALSE), nrow(linebacker_data), replace=TRUE, prob=c(0.8,0.2))
data_LBs_train <- data_LBs[train_indices, ]
data_LBs_test <- data_LBs[-train_indices, ]

# Prepare data
X_train <- data_LBs_train[, selected_variables]
y_train <- data_LBs_train$Time_Spent_Engaged
X_test <- data_LBs_test[, selected_variables]
y_test <- data_LBs_test$Time_Spent_Engaged

# Convert data to DMatrix format
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
dtest <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)

library(xgboost)

# Set seed for reproducibility
set.seed(1842)

# Convert data to DMatrix format
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
dtest <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)

# Set XGBoost parameters
params <- list(
  objective = "reg:linear",
  eval_metric = "rmse",
  seed = 1842
)

# Train the XGBoost model with early stopping
xgb_model <- xgb.train(
  data = dtrain,
  params = params,
  nrounds = 1000,
  watchlist = list(train = dtrain, test = dtest),
  early_stopping_rounds = 10,
  verbose = 1
)

# Output the best iteration
best_iteration <- xgb_model$best_iteration
cat("Best Iteration:", best_iteration, "\n")

# Use the XGBoost model for predictions
y_pred <- predict(xgb_model, newdata = as.matrix(X_test))

# Evaluate the model on the testing set
rmse <- sqrt(mean((y_test - y_pred)^2))
cat("Root Mean Squared Error (RMSE) on Testing Set:", rmse, "\n")
```


# Now let's make predictions with our model:
```{r}
# Use the XGBoost model for predictions
predictions <- predict(xgb_model, as.matrix(X))

# Assuming you have a data frame named data_LBs
linebacker_predictions <- data.frame(Linebacker_Name = data_LBs$Linebacker_Name, Predicted_Time = predictions)

# Assuming your dataset is named "predictions_data"
library(dplyr)

# Convert "Predicted_Time" to numeric if it's not already
linebacker_predictions$Predicted_Time <- as.numeric(linebacker_predictions$Predicted_Time)

# Calculate the average predicted time for each linebacker
average_times <- linebacker_predictions %>%
  group_by(Linebacker_Name) %>%
  summarise(Average_Predicted_Time = mean(Predicted_Time))

# Order the results by ascending average predicted time
ranked_LBs <- average_times[order(average_times$Average_Predicted_Time), ]

# Print the ordered results
print(ranked_LBs, n = 100)
```



```{r}
# Load necessary libraries
library(xgboost)
library(ggplot2)

# Assuming you have already trained an XGBoost model named xgb_model
# If not, you should train your model first

# Extract feature importance scores
importance_matrix <- xgb.importance(model = xgb_model)

# Convert importance matrix to a data.frame
importance_df <- as.data.frame(importance_matrix)

# Order the data frame by feature importance in descending order
importance_df <- importance_df[order(-importance_df$Gain),]

# Plot feature importance with reversed x-axis and annotate bars with values
ggplot(importance_df, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(Gain, 6)), vjust = .5, hjust = 1, size = 7, color = "white") +  # Add text annotations
  labs(title = "XGBoost Feature Importance",
       x = "Feature",
       y = "Importance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()
```


## Display the Top 10 Linebackers Based on In-Play Data Feature Importance 
```{r}
# Load necessary libraries
library(ggplot2)
library(ggimage)

# Set your custom colors
custom_colors <- c("#FB4F14", "#D50A0A","#0B2265", "#125740", "#A71930", "#008E97","#006778","#5A1414", "#FF3C00", "#69BE28","#FB4F14","#AA0000","#D50A0A","#E31837","#241773", "#008E97", "#0B2265", "#125740","#004C54","#241773","#0085CA", "#D50A0A", "#0B2265","#C60C30","#00338D")


# Set your image folder path
image_folder <- "player_images3"

# Combine image file names with full path
image_files <- c("Von_Miller.png", "Zaire_Franklin.png", "Kwon_Alexander.png", "Roquan_Smith.png", "Robert_Quinn.png", "Alex_Singleton.png", "Shaquil_Barrett.png", "Melvin_Ingram.png", "Anthony_Nelson.png", "T.J._Edwards.png", "Lavonte_David.png", "Shaq_Thompson.png", "Devin_Loyd.png", "Jeremiah_Owosu.png", "Oshane_Ximines.png", "David_Long.png", "Leo_Chenal.png", "Josey_Jewell.png", "Malik_Harrison.png", "Jamin_Davis.png", "Quincy_Williams.png", "C.J._Mosley.png", "Jordyn_Brooks.png", "Bobby_Okereke.png", "Christian_Harris.png")
image_files <- paste0(image_folder, "/", image_files)

# Create a data frame with image information
image_data <- data.frame(x = rep(Inf, length(image_files)),
                         y = rep(0, length(image_files)),
                         image = image_files,
                         Linebacker_Name = c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris"),
                         Seconds = c(0.721, 0.744, 0.762, 0.764, 0.765, 
                                             0.766, 0.770, 0.793, 0.794, 0.796, 
                                             0.802, 0.806, 0.806, 0.813, 0.818, 
                                             0.820, 0.823, 0.823, 0.825, 0.828, 
                                             0.831, 0.833, 0.834, 0.838, 0.844),
                         Adjustment = c(-0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002, -0.002),
                         stringsAsFactors = FALSE)

# Add a new column for image size
image_data$size <- ifelse(image_data$Linebacker_Name %in% c("Cody"), 0.12, 0.06)

# Add a new column for image size
image_data$color <- "black"

# Add a new column for text size
image_data$size2 <- 7

# Add a new column for image size
image_data$nudge <- 0.4

ggplot(image_data, aes(x = reorder(Linebacker_Name, desc(Seconds)), y = Seconds)) +
  geom_bar(stat = "identity", aes(fill = Linebacker_Name), position = position_dodge(width = 1.2), size = 2) +
  geom_image(aes(x = reorder(Linebacker_Name, desc(Seconds)), y = Seconds + Adjustment, image = image), size = image_data$size, by = "height", position = position_nudge(x = image_data$nudge)) +
  geom_text(aes(x = reorder(Linebacker_Name, desc(Seconds)), y = Seconds, 
                label = sprintf("%.4f", Seconds)),
            vjust = 0.2, hjust = -0.3, size = image_data$size2, color = "black") +  
  scale_fill_manual(values = custom_colors) +
  labs(title = "Top 25 Players - Play Metrics Predicted Seconds",
       x = "Linebacker Name",
       y = "Predicted Seconds",
       title_font = "Helvetica",
       title_face = "bold",
       title_size = 20,
       x_font = "Helvetica",
       x_face = "bold",
       x_size = 12,
       y_font = "Helvetica",
       y_face = "bold",
       y_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 10),
        axis.title.y = element_text(face = "bold", size = 15),
        axis.title.x = element_text(face = "bold", size = 15),
        axis.text.y = element_text(face = "bold", size = 10),
        legend.position = "none") +
  coord_flip(ylim = c(min(image_data$Seconds) - 0.06, max(image_data$Seconds) + 0.1), 
             xlim = c(0, 26))
```



## Display the Top 10 Linebackers Based on Time Spent on Block 
```{r}
# Add Average_Seconds to dataframe
data_LBs <- data_LBs %>%
  group_by(Linebacker_Name) %>%
  mutate(Average_Seconds = mean(Time_Spent_Engaged, na.rm = TRUE))

bottom_100_lowest_seconds <- data_LBs %>%
  group_by(Linebacker_Name) %>%
  summarize(Average_Seconds = sprintf("%.4f", mean(Time_Spent_Engaged, na.rm = TRUE))) %>%
  arrange(Average_Seconds) %>%
  slice_head(n = 100)

# Create Linebacker_Tackle_Total variable
data_LBs <- data_LBs %>%
  distinct(gameId, playId, Linebacker_Name, .keep_all = TRUE) %>%
  group_by(Linebacker_Name) %>%
  mutate(Linebacker_Tackle_Total = sum(tackle, na.rm = TRUE) + sum(assist, na.rm = TRUE)) %>%
  ungroup()

# View the resulting dataset
print(data_LBs$Linebacker_Tackle_Total)

# Display the bottom 10 linebackers with the lowest average seconds
print(bottom_100_lowest_seconds, n = 100)


# Set your image folder path
image_folder <- "player_images3"

# Combine image file names with full path
image_files <- c("Zaire_Franklin.png", "Roquan_Smith.png", "Kwon_Alexander.png", "Robert_Quinn.png", "Malik_Harrison.png", "Melvin_Ingram.png", "T.J._Edwards.png", "Josey_Jewell.png", "Shaq_Thompson.png", "Von_Miller.png", "Bobby_Okereke.png", "Alex_Singleton.png", "C.J._Mosley.png", "Leighton_Vander_Esch.png", "Shaquil_Barrett.png", "Anthony_Nelson.png", "Matt_Milano.png", "Oshane_Ximines.png", "Kyzir_White.png", "Bobby_Wagner.png", "Lavonte_David.png", "Quincy_Williams.png", "Jordyn_Brooks.png", "David_Long.png", "Tae_Crowder.png")
image_files <- paste0(image_folder, "/", image_files)

# Create a data frame with image information
image_data <- data.frame(x = rep(Inf, length(image_files)),
                         y = rep(0, length(image_files)),
                         image = image_files,
                         Linebacker_Name = c("Zaire Franklin", "Roquan Smith", "Kwon Alexander", "Robert Quinn", "Malik Harrison", "Melvin Ingram", "T.J. Edwards", "Josey Jewell", "Shaq Thompson", "Von Miller", "Bobby Okereke", "Alex Singleton", "C.J. Mosley", "Leighton Vander Esch", "Shaquil Barrett", "Anthony Nelson", "Matt Milano", "Oshane Ximines", "Kyzir White", "Bobby Wagner", "Lavonte David", "Quincy Williams", "Jordyn Brooks", "David Long", "Tae Crowder"),
                         Average_Seconds = c(0.7114, 0.7289, 0.7390, 0.7448, 0.7657, 0.7914, 0.7932, 0.8000, 0.8010, 0.8025, 0.8071, 0.8111, 0.8140, 0.8230, 0.8300, 0.8302, 0.8352, 0.8367, 0.8385, 0.8403, 0.8418, 0.8460, 0.8469, 0.8509, 0.8538),
                         Adjustment = c(-0.04, -0.035, -0.035, -0.025, -0.03, -0.03, -0.05, -0.03, -0.015, -0.05, -0.015, -0.002, -0.035, -0.002, -0.045, -0.03, -0.002, -0.03, -0.002, -0.002, -0.03, -0.05, -0.015, -0.045, -0.002),
                         stringsAsFactors = FALSE)

# Add a new column for image size
image_data$size <- ifelse(image_data$Linebacker_Name %in% c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris"), 0.12, 0.06)

# Add a new column for image size
image_data$size2 <- ifelse(image_data$Linebacker_Name %in% c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris"), 9, 8)

# Add a new column for image size
image_data$nudge <- ifelse(image_data$Linebacker_Name %in% c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris"), 1.2, 0.4)

# Add a new column for image size
image_data$color <- ifelse(image_data$Linebacker_Name %in% c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris"), "white", "darkblue")

# Add a new column for image size
image_data$color2 <- ifelse(image_data$Linebacker_Name %in% c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris"), "#FBF719", NA)

# Set your custom colors
custom_colors <- c("#FB4F14","#FF7900","#0B2265","#69BE28", "#125740","#008E97", "#69BE28", "#FB4F14","#AA0000","#C41E3A", "#FF7900","#003594","#241773","#C60C30", "#008E97","#241773", "#125740", "#125740","#241773","#0085CA","#FF7900", "#C83803", "#0B2265","#C60C30","#002C5F")

highlight_names <- c("Von Miller", "Zaire Franklin", "Kwon Alexander", "Roquan Smith", "Robert Quinn", "Alex Singleton", "Shaquil Barrett", "Melvin Ingram", "Anthony Nelson", "T.J. Edwards", "Lavonte David", "Shaq Thompson", "Devin Loyd", "Jeremiah Owosu", "Oshane Ximines", "David Long", "Leo Chenal", "Josey Jewell", "Malik Harrison", "Jamin Davis", "Quincy Williams", "C.J. Mosley", "Jordyn Brooks", "Bobby Okereke", "Christian Harris")

ggplot(image_data, aes(x = reorder(Linebacker_Name, desc(Average_Seconds)), y = Average_Seconds)) +
  geom_bar(stat = "identity", aes(fill = Linebacker_Name), size = 2.5) +
  geom_image(aes(x = reorder(Linebacker_Name, desc(Average_Seconds)), y = Average_Seconds + Adjustment, image = image), size = image_data$size, by = "height", position = position_nudge(x = image_data$nudge)) +
  geom_text(aes(x = reorder(Linebacker_Name, desc(Average_Seconds)), y = Average_Seconds, 
                label = sprintf("%.4f", Average_Seconds), color = color),
            vjust = .5, hjust = -0.2, size = image_data$size2) +  
  scale_fill_manual(values = custom_colors) +
  labs(title = "Top 25 Players - Time Spent on Block",
       x = "Linebacker Name",
       y = "Seconds",
       title_font = "Helvetica",
       title_face = "bold",
       title_size = 25,
       x_font = "Helvetica",
       x_face = "bold",
       x_size = 10,
       y_font = "Helvetica",
       y_face = "bold",
       y_size = 15) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.title.y = element_text(face = "bold", size = 15),
        axis.title.x = element_text(face = "bold", size = 15),
        axis.text.y = element_text(face = "bold", size = 10),
        legend.position = "none") +
  coord_flip(ylim = c(min(image_data$Average_Seconds) - 0.06, max(image_data$Average_Seconds) + 0.25), 
             xlim = c(0.2, 27))
```

```{r}
library(dplyr)

# Load the dataset
data <- read.csv("BDB_all_weeks.csv") 

# Load the tackles dataset
tackles <- read.csv("tackles.csv")

# Assuming 'tackles' and 'BDB_all_weeks' are your datasets

result <- tackles %>%
  left_join(data, by = c("gameId", "playId", "nflId")) %>%
  group_by(gameId, playId, nflId)

# Assuming 'your_data' is your dataset
result <- result %>%
  distinct(gameId, playId, .keep_all = TRUE)

result <- result %>%
  group_by(nflId) %>%
  summarise(nflId = nflId, total_tackles = sum(tackle) + sum(assist))
result

result <- result %>% distinct()
result

# Extract 'displayName' column from 'your_data'
display_names <- unique(data$displayName)
print(display_names)

# Filter to only linebackers
data2 <- data %>% filter(final_position %in% c("MLB", "OLB", "ILB"))
unique(data2$displayName)

joined_data <- left_join(result, data2, by = "nflId")

selected_data <- joined_data %>% 
  select(gameId, playId, nflId, displayName, total_tackles)
print(selected_data)

selected_data <- na.omit(selected_data)

selected_data <- selected_data %>% 
  group_by(displayName) %>% 
  mutate(total_plays = n_distinct(paste(gameId, playId)))

unique_data <- selected_data %>% 
  distinct(displayName, .keep_all = TRUE)

unique_data$tackles_per_play <- unique_data$total_tackles/unique_data$total_plays

top_100_tacklers <- unique_data %>% 
  arrange(desc(total_tackles))

top_100_tacklers <- top_100_tacklers %>% 
  filter(total_plays >= 30)

print(top_100_tacklers, n = 229)

library(dplyr)

# Assuming df1 is the first data frame and df2 is the second data frame
# First, change "Linebacker_Name" to "displayName" in the second data frame
bottom_100_lowest_seconds <- bottom_100_lowest_seconds %>% rename(displayName = Linebacker_Name)

# Then, left join the two data frames on the common column "displayName"
result_df <- top_100_tacklers %>% left_join(bottom_100_lowest_seconds, by = "displayName")

# Print the resulting data frame
print(result_df, n = 100)

```

